---
title: "spatial_tutorial"
output: html_document
date: "2024-01-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# visualization
library(mapview)

# data exploration and manipulation
library(tidyverse) # suite of packages for data manipulation and exploration
library(here) # makes it easy to call files despite differences in file pathways

# spatial analysis
library(sf) # sf = simple feature, the package to use to do GIS in R
library(gstat) # does spatial stats
library(stars) # one method of opening and using raster data
library(terra) # another method of opening and using raster data
library(tidyterra) # allows for tidyverse and ggplot manipulations of terra objects
library(cubelyr)
```

we have lots of data on soil salinity, in different forms, some explicitly spatial and some not

sf = simple feature (spatial)

```{r read in data}
soils <- read_csv(here("data", "soil_traits.csv"))

locations_sf <- read_sf(here("data", "sample_locations.shp")) %>% 
  st_set_crs(value = 4326) # proper geo projection is WGS 1984 - lat/lot as degrees, min, sec
# geometry is a sticky item - sticks no matter how you manipulate the data

# join spatial information to nonspatial

soils_sf <- locations_sf %>% 
  merge(soils) # merges by commonly names columns by default, recommended for this to avoid multiple columns showing up in final dataset


```
# raster data
## stars methods
```{r raster data with stars method}
# alternative to raster package, not a standalone replacement, but joined in with sf can do almost everything you would have done with raster

# purely for raster data

dem_stars <- read_stars(here("data", "dem.tif")) #%>%  #type into console to view dimensions, offset (starting point), delta = spatial resolition, attribute (ours is dem.tif)
# if you wanted to check ??or change?? coord system
  #st_warp(crs = 4236)



# visualize
ggplot()+
  geom_stars(data = dem_stars, aes(x = x, y = y))


# might want to crop it using a boundary shapefile (that alreday exists?)

bbox <- read_sf(here("data", "bbox.shp")) # where did this shapefile come from?
# in arc, made a bounding box and saved as a file
# if you know the coords of your corners, can make a csv and save as a polygon in sf

dem_stars <- dem_stars %>% 
  st_crop(y = st_bbox(bbox)) #st = sf package, st_bbox is command to find boundary of object (which we caleld bbox)
  
```


## terra methods
```{r raster data with terra method}
# terra was meant to replace raster fully
# has native options for both raster data and vector data

dem_terra <- rast(here("data", "dem.tif")) # terra option for opening raster file

# saves it as SpatRaster class file

# cropping
bbox_terra <- vect(here("data", "bbox.shp")) # vect is how to open shapefiles

# has it's own native cropping
terra_crop <- terra::crop(dem_terra, bbox_terra)

# to use terra objects in ggplot (or tidyverse), you need to have tidyterra installed!

ggplot()+
  geom_spatraster(data = terra_crop)
```

# rest of the tutorial will be in stars
## opening a few more raster layers, like veg indices
```{r some other rasters}

ndvi_stars <- read_stars(here("data", "2022_03_22_ndvi.tif")) %>% 
  st_warp(crs = 4326) # checking projection (not defining projection if it already has one)

mari_stars <- read_stars(here("data", "03_22_mari_avg.tif")) %>%  # red pigments in plants, stress that they have
  st_warp(crs = 4326)
```

crs is how we mathematically make a globe into a flat shape, different projections tell us what math we used to preserve the shape of the earth when translating into a flat surface
eg wgs84 ensures that we still have lat/lot rather than a meter based coordinate, most gps data will be collected in this

## helpful to merge all of our raster files together
```{r merging raster files}

#rast_merge <- c(ndvi_stars, mari_stars, dem_stars) # won't work, we have to specify a parameter (spatial resolutions are different so it doesn't know how to align the data sets)

# we can use stwarp
dem_warp <- dem_stars %>% 
  st_warp(dest = ndvi_stars) # tell it what you want the desination of the warping to be, tell it to match ndvi
#will reproject and regrid!

rast_merge <- c(ndvi_stars, mari_stars, dem_warp) # replace dem_stars with new dem_warp

rast_merge

```

# lets fuck with the data

```{r manipulations of raster_merge using sf and tidyverse}
# rename raster attributions to be easier to type using raster_merge

rast_r <- rast_merge %>% 
  setNames(c("NDVI", "mARI", "Elevation")) #%>% 
 # plot # will plot out first attribute

plot(rast_r[,,2]) # not working but should??

# what if you didn't want to use files separate and wanted them to be another dimension on your dataset?

rast_m <- rast_r %>% 
  merge()

# now we only have one attribute that has the combined names, but if we look at dimension, we have all three attributions
# now we have layers, used ot be raster stack

plot(rast_m) # now we can see all three attributes
# can only filter when it's merged like this, can't do it when it's separate because there's no way to call one particular attribute
# eg for a range of ndvi's

rast_s <- rast_m %>% 
  split() # how we can pull the attributes apart again

plot(rast_m[,,,2]) #plot 2nd attribute

# filter now (only based on dimensions, not attributes)
# we want only less than 34.417

rast_f <- rast_r %>% 
  filter(y < 34.417)

plot(rast_f)

# can specify which attribute to plot with brackets

```
# next we will extract rasters

```{r extraction of rasters}

rast_extract <- rast_r %>% 
  st_extract(soils_sf) %>% 
  st_drop_geometry()  # want to avoid sticky redundancy with geometry layer to need to convert to plain dataframe without duplicate geometries

# now we will join with soils simple feature

soils_rast <- soils_sf %>% 
  bind_cols(rast_extract) # different way to combine things when there is no default matching col names
```

# lets make some simple maps with ggplot

```{r maps!}
ggplot()+ # have to put things in layers from most area covered to least
  geom_stars(data = dem_stars, #base map
             aes(x = x, y = y,  fill = dem.tif),
             na.action = na.omit) +
  scale_fill_gradientn(colors = c("white", "bisque4", "black")) # don't forget the n!
  #geom_sf(data = soils_rast, # point data
   #       aes(color = electro_cond_mS_per_cm, shape = landcover)) #don't have to specify x and y because geom sf looks for it


# fully interactive map!
mapview(dem_stars)+ #works in rstudio and html files
  mapview(soils_rast)

```

