---
title: "Manipulating and Using Spatial Data"
output: html_document
date: "2024-01-11"
---

Required packages: Copy and paste in console as necessary
install.packages("here")
install.packages("sf")
install.packages("mapview")
install.packages("gstat")
install.packages("stars")
install.packages("terra")
install.packages("tidyterra")


```{r setup, include=FALSE, message=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)

# visualization
library(mapview)

# data exploration and manipulation
library(tidyverse) # suite of packages for data manipulation and exploration
library(here) # makes it easy to call files despite differences in file pathways

# spatial analysis
library(sf) # sf = simple feature, the package to use to do GIS in R
library(gstat) # does spatial stats
library(stars) # one method of opening and using raster data
library(terra) # another method of opening and using raster data
library(tidyterra) # allows for tidyverse and ggplot manipulations of terra objects
```

new chunk = ctrl + alt + i / cmd+option+i
pipe = cmd + shft + m / ctrl + shft + m

# Opening and exploring data

## Open soil data (spatial and aspatial)

Spatial data comes to us as vector (points, lines, polygons) or raster (continuous grid) data and has a location or geographic reference associated with the files (location), aspatial data is any data that does not have a spatial component (country name, plant name, leaf size, etc).

```{r}
#opening soil trait csv
soils <- read_csv(here("data", "soil_traits.csv")) # opening your soil trait data (aspatial data)

# open sampling location spatial data
locations_sf <- read_sf(here("data", "sample_locations.shp")) %>% 
  st_set_crs(value = 4326)

# or can open if you have a csv with lat/lon coordinates and you know the projection of the data
locations <- read_csv(here("data", "sample_locations.csv")) %>% # open the csv file
  st_as_sf(coords = c("longitude", "latitude")) %>% # tell code to read this file as a sf dataset using the "longitude" and "latitude" columns as coordinates
  st_set_crs(value = 4326) # set the data projection to coordinate reference system code 4326 (WGS 1984)

# joining the data files so that you have traits alongside spatial location
soils_sf <- locations_sf %>% # call the soil trait data
  merge(soils) # merge the data.frame style objects together (by default uses common column names)
```

## Open Elevation Raster Data

### Method 1: `{stars}` package

```{r}
# Opening the Digital Elevation Model file
dem_stars <- read_stars(here("data", "dem.tif")) %>% # open and read a stars object
  st_warp(crs = 4326) # set the coordinate reference system

# call the stars raster object to see how R stores the data (will contrast to {terra})

# plot the file to see how large it is
ggplot()+
  geom_stars(data = dem_stars, aes(x = x, y = y)) # geom_stars() works to visualize stars objects

# opening vector for cropping data
bbox <- read_sf(here("data", "bbox.shp")) #read in bounding box file (or the boundary used to clip the raster data)

# Map both files to show how much will be cropped out
ggplot()+
  geom_stars(data = dem_stars, aes(x = x, y = y)) + # geom_stars() works to visualize stars objects
  geom_sf(data = bbox, fill = NA, color = "black", linewidth = 2) # geom_sf lets us use vector files as 

## cropping data
dem_stars <- read_stars(here("data", "dem.tif")) %>% # read in dem 
  st_crop(y = st_bbox(bbox))  #crop to bounding box

## plotting cropped raster
ggplot()+
  geom_stars(data = dem_stars, aes(x = x, y = y)) + # geom_stars() works to visualize stars objects
  coord_sf(crs = 4326) # sets the aspect ratio of the map to follow spatial geometry (aka no default stretching)
```


### Method 2: `{terra}` package
```{r}
dem_terra <- rast(here("data", "dem.tif")) # open a SpatRast object

# call terra SpatRast object to compare to the stars object

#crop to the same area as stars

bbox_vect <- vect(here("data", "bbox.shp")) # open bounding box as SpatVect object

terra_crop <- terra::crop(dem_terra, bbox_vect) # crop raster to boundary

# plot the resulting raster
ggplot()+
  geom_spatraster(data = terra_crop) #geom_spatraster comes from tidyterra (notice no need for setting aspect ratio)
```

Based on my preference and experience with data manipulations with one package over the other, the rest of the tutorial uses `{stars}`

## Let's open a few other raster layers

```{r}
#ndvi
ndvi_stars <- read_stars(here("data", "2022_03_22_ndvi.tif")) %>% # open a normalized difference vegetation index file 
  st_warp(crs = 4326)

#mARI
mari_stars <- read_stars(here("data", "03_22_mari_avg.tif")) %>% # open a modified anthocyanin reflectance index file
  st_warp(crs = 4326)
```

# Merging and Manipulating Raster Data

## Merging mulitple raster files

```{r}
# merge the raster files using `c()`
#rast_merge <- c(ndvi_stars, mari_stars, dem_stars)

## we get an error? why? examine the files ndvi_stars and dem_stars

## make the dimensions match between the files using `st_warp()` (a spatial resample/transform function)
dem_warp <- dem_stars %>% # file we want to transform
  st_warp(dest = ndvi_stars) # target format

## call to make sure it worked

## attempt merge again
rast_all <- c(ndvi_stars, mari_stars, dem_warp)
```

## Manipulating `{stars}` using `{tidyverse}` and `{sf}`

 Many tidyverse and sf functions work with stars objects here are a few examples of what can be done
 
 More Examples with explanations can be found here: https://tmieno2.github.io/R-as-GIS-for-Economists/stars-basics.html

```{r}
# renaming attributes
rast_r <- rast_all %>% 
  setNames(c("NDVI", "mARI", "Elevation"))

# merging attributes into dimensions
rast_merge <- rast_r %>% 
  merge()

# spliting dimensions
rast_sp <- rast_merge %>% 
  split()

# filter (dimensions)
rast_f <- rast_r %>% 
  filter(y < 34.417)

# select (attributes)
rast_sl <- rast_r %>% 
  select(NDVI, Elevation)

# reprojecting and transforming data
rast_w <- rast_r %>% 
  st_warp(crs = 32611)

# plotting single or all attributes
# By default only first attribute gets plotted with plot()
plot(rast_r)
plot(rast_r["mARI",,])
plot(rast_r["Elevation",,])

# or you can added plot() at the end of a data manipulation to quickly visualize the changes
rast_merge <- rast_r %>% 
  merge() %>% 
  plot()

rast_f <- rast_r %>% 
  filter(y < 34.417) %>% 
  plot()
```


# Extracting Raster Data

```{r}
rast_extract <- rast_r %>%  # call renamed raster file
  st_extract(soils_sf) %>% # extract data based on point location
  st_drop_geometry() # convert from sf to data frame

soils_rast <- soils_sf %>% # call sf file
  bind_cols(rast_extract) # add extracted columns to sf file

# view data
soils_rast
```

# Making Maps

## `ggplot()` maps 

```{r}
ggplot() +
  geom_stars(data = dem_stars, aes(x = x, y = y, fill = dem.tif), na.action = na.omit) +
  scale_fill_gradientn(colors = c("white", "bisque4", "black")) +
  geom_sf(data = soils_rast, aes(color = electro_cond_mS_per_cm, shape = landcover))
  

# Adjust the scale if you'd like
ggplot() +
  geom_stars(data = dem_stars, aes(x = x, y = y, fill = dem.tif), na.action = na.omit) +
  scale_fill_gradientn(colors = c("white", "bisque4", "black")) +
  scale_y_continuous(limits = c(34.411, 34.417)) +
  scale_x_continuous(limits = c(-119.878, -119.873))+
  geom_sf(data = soils_rast, aes(color = electro_cond_mS_per_cm, shape = landcover))
```

## Interactive maps w/ `mapview()` (works in RStudio and HTML files)

```{r}
mapview(dem_stars) + 
  mapview(soils_rast)
```


